Working with daily climate model output data in R and the \pkg{futureheatwaves} package

### 2017-01-04 paper_drafts/RJournalRevisions/anderson-eason-barnes.tex
More work on article draft

\abstract{%
Research on climate change impacts can require extensive processing to
incorporate output from ensembles of different climate models and
different simulations of each model. This processing can be particularly
extensive when identifying and characterizing extreme events like heat
waves and frost day spells, as these must be processed from model output
with daily time steps and may use identification threholds based on
local distributions (e.g., 98\textsuperscript{th} percentile temperature
for that location). Here, we provide an overview of working with daily
climate model output data in R. Further, we present the
\pkg{futureheatwaves} package, which we developed to ease the process of
identifying, characterizing, and exploring multi-day extreme events in
climate model output. This package takes as input a directory of
projection files from one or many climate models, identifies all
occurrences of a specific type of extreme event (e.g., heat waves) for
multiple cities using a user-specified event definition, and writes an
output directory of files with these extreme events and their
characteristics (e.g., length, absolute and relative temperatures). The
\pkg{futureheatwaves} package then allows the user to create custom
functions and apply them across all event files; this functionality can
be used either to explore extreme event patterns (e.g., average length
and intensity across different climate models) or to apply
epidemiological estimates to project health impacts under different
scenarios.
}

@@

Research on climate change impacts can require extensive processing.
This is not only because output files for a single climate model can be
large, but also because of the rising popularity of ensemble techniques
\citep{IPCCch1}, in which, to better characterize uncertainty in
projections, impacts are assessed for multiple climate models, multiple
simulations of each climate model, and multiple climate experiments.
Projections of regional climate change over the next century are subject
to high uncertainty due to three distinct sources: (1) internal climate
variability, i.e.~climate noise, (2) climate model uncertainty, i.e.~the
same forcing can produce a different response in different models and
(3) scenario uncertainty, i.e.~uncertainty in future climate forcings
(e.g. \citet{hawkins2009potential}). One approach for ensuring that
these sources of uncertainty are characterized is to simulate the future
climate many times with multiple models and for multiple future
scenarios.

For example, the Coupled Model Intercomparison Project, phase 5 (CMIP5;
\citet{taylor2012overview}) brought together dozens of major climate
modeling groups around the world to simulate the same future radiative
forcing scenarios, but with their own models. This created an ensemble
of state-of-the-art climate model projections that allow researchers to
study projections and their uncertainties. Most of these modeling groups
additionally performed more than one simulation for each scenario and
model (i.e.~multiple ensemble members), perturbing the initial
conditions by a very tiny amount to quantify uncertainties due to
internal climate variability.

Processing climate output data is particularly intensive for data in
smaller time steps, like daily output. While some climate impacts can be
assessed using climate model output at monthly, seasonal, or yearly time
steps, the impacts of multi-day extreme events must be assessed using
output in daily time steps. Such multi-day extreme events include heat
waves, cold spells, frost day spells, and droughts. The assessment is
further complicated if extreme events are identified based on conditions
that are rare for a certain location (e.g., 98th percentile of local
temperature distribution for identifying heat waves) \citep{IPCCch1}, as
this event definition must be determined at each study location from
climate model output. Further, it is often of interest to create
summaries of multiple characteristics of these extreme events. For
example, one may be interested in determining whether the frequency or
characteristics (e.g., length, intensity) of heat waves or warm spells
will change under certain climate change scenarios \citep{IPCCch1}.

We begin this article with an overview of climate model output data,
particularly daily data, for R users. We outline where data from CMIP5
can be obtained as well as how to work with the file format (netCDF)
from R. We overview some R packages that can be useful when working with
this data, as well as aspects of the data (e.g., non-standard calendars)
of which users should be aware when working with daily climate model
output in R.

After this overview, we present the \pkg{futureheatwaves} package which
we created to aid in identifying and characterizing any type of
multi-day extreme event from daily climate model output (Table
\ref{tab:goals}). This package can be used to identify and characterize
a variety of multi-day extreme events, but provides some functionality
particularly useful in identifying and characterizing heat waves
specifically. Quantification of the impacts of heat waves on human
health suffer from additional sources of uncertainty beyond those
inherent in projections of regional changes in surface temperature.
These include: (1) uncertainty in the ability of communities to adapt to
changing temperatures (the adaptation scenario) and (2) uncertainty in
the definition of a heat wave itself. Thus, identifying and
characterizing the impacts of future heat waves in state-of-the-art
climate models requires analyzing hundreds of projections based on
combinations of anthropogenic activity, climate model and ensemble
member, adaptation scenario, and heat wave definition. Such an analysis
is non-trivial. For example, to estimate the possible impacts of heat
waves using 32 ensemble members of CMIP5, two forcing scenarios, two
different heat wave definitions, and five progressive assumptions about
adaptation to heat, would require that one identify and characterize
heat waves in 640 separate temperature time series for each region of
interest.

\begin{table}[t]
\begin{center}
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\begin{tabular}{lL{\dimexpr0.8\textwidth-2\tabcolsep\relax}}
\toprule
 & Design goals of the \pkg{futureheatwaves} package \\
\midrule
1 & Make processing of large sets of climate projections more practical for researchers exploring the potential impacts of heat waves and other multi-day extreme events.\\
2 & Speed up processing time by incorporating C++ in event identification.\\
3 & Keep track of the names of climate models, and number of ensemble members processed for each.\\
4 & Not only identify, but also characterize, all extreme events within each climate projection, to allow the exploration of patterns in these characteristics across different projections and also to allow the use of more complex impacts models, including models that include effect modification by event characteristics (e.g., event length, event intensity). For example, this package allows the user to apply a health effects model where risk of mortality is not the same for every heat wave, but rather is modified by heat wave length, intensity, or other measured characteristics.\\
5 & Give users extensive power in customizing the process, including allowing custom event definitions.\\
6 & Allow users to easily explore the extreme events identified within all climate projections by applying custom functions across heat wave data sets from all projections at once.\\
7 & Create output that is in a ``tidy" data format, allowing it to work well with \pkg{ggplot2} for visualization.\\
\bottomrule
\end{tabular}
\end{center}
\caption{Design goals for the \pkg{futureheatwaves} package.}
\label{tab:goals}
\end{table}

The \pkg{futureheatwaves} package automates the process of identifying
and characterizing multi-day extreme events across different ensemble
members of one or more climate models. A variety of different heat wave
definitions have been used to identify heat waves in a time series of
temperature data \citep{smith2013heat}, and the choice of heat wave
definitions can influence both health effect estimates
\citep{chen2015influence, kent2014heat} and projected heat wave trends
\citep{smith2013heat}. Further, other types of extreme events will be
defined differently than heat waves (for example, frost day spells may
be defined as one or more days with temperature at or below
32\textsuperscript{o}F / 0\textsuperscript{o}C). This package therefore
allows the user to create and use a custom extreme event definition to
identify events in the climate model output. Finally, this package
allows the user to explore the extreme events identified in the climate
model output with a function that can take a user-defined R function and
apply it across all extreme event files generated for the separate
ensemble members.

@@

For climate impact studies, a main source for climate model output is
the Coupled Model Intercomparison Project, which is currently in its
fifth phase (CMIP5). Over 20 climate modeling groups have created one or
more climate models which, for this project, are run using standardized
scenarios \citep{taylor2012overview}. The resulting output is uniform
across modeling groups and has a consistent structure, which allows
comparison of simulations from different models \citep{IPCCch9}. For
CMIP5, each group ran simulations under several experiments, with
experiments varying in terms of the radiative forcing. This radiative
forcing depends on time-varying model inputs (greenhouse gas emissions
or concentrations, land use changes, etc.), which are specified for each
experiment \citep{taylor2012overview, IPCCch9}. Experiments include
historical experiments (run using forcings consistent with observed and
reconstructed data for 1850--2005), pre-Industrial control experiments,
and experiments of future scenarios of radiative forcing over the 21st
century or longer (e.g., RCP4.5, RCP8.5) \citep{taylor2012overview}.
Some modeling groups created ensembles of output for a specific model
and experiment, in which they ran the experiment multiple times with the
model with very small changes to the initial conditions, resulting in
multiple ensemble members for a single climate model and experiment.
CMIP5 climate model output is archived at a number of different time
steps (e.g., daily, monthly, seasonal, yearly) \citep{taylor2010cmip5},
and some variables are reported at multiple levels in the ocean or
atmosphere (e.g., ocean temperature). Here, we will focus on data with a
daily time step for variables reported at a single level (e.g.,
near-surface air temperature).

The CMIP5 climate model output data is distributed across data nodes at
different climate modeling centers \citep{taylor2012overview}, but can
be accessed centrally at the World Climate Research Programme CMIP5 data
portal at \url{https://pcmdi.llnl.gov/search/cmip5/}. Users must
register before downloading data, and some data are restricted to
non-commercial use. There is a separate file for each combination of
climate model, experiment, modeling realm (e.g., atmosphere, ocean),
variable, time step, and ensemble member
\citep{taylor2012overview, taylor2010cmip5}. For finer time scales, this
output is further split across multiple files for specific year ranges
(e.g., 5 years of output for each file) \citep{taylor2010cmip5}.
Filenames for CMIP5 files can be parsed to generate information about
the output variable, climate model, experiment, and ensemble member for
the simulation \citep{taylor2010cmip5}. Files can be searched and
downloaded through a point-and-click web interface. The data portal also
allows you to generate a specific wget script, which you can use to
download many files at once. A wget script can also be generated through
Earth System Grid Federation's Search RESTful API. Tips on efficiently
searching and downloading the data, including through use of wget
scripts and the search API, are available as user tutorials through the
website of the University of Colorado Boulder's Earth System CoG (e.g.,
\url{https://www.earthsystemcog.org/projects/cog/doc/wget} for a
tutorial on downloading files using wget).

CMIP5 files are saved in Network Common Data Format (netCDF). NetCDF is
a binary file format that allows storage of data representing a regular
array. These climate model output files can be as large as several
gigabytes {[}?{]}. For climate model output at a single level (e.g.,
near-surface air temperature), the data is a 3-dimensional array, with
dimensions representing time and two coordinates of location (e.g.,
latitude and longitude) (Figure \ref{fig:netcdfexample}). Global climate
models generate output at regularly-spaced time steps, typically at
regularly-spaced grid points around the world. The latitude and
longitude spacing of grid points vary by climate model, but are
typically 1--2 degrees for atmospheric variables in CMIP5 models
\citep{IPCCch9}. Each data point in the netCDF array gives the modeled
value of the variable (e.g., surface temperature) for a single time
point and location. For CMIP5 climate model output, the location units
are in degrees east and degrees north for longitude and latitude,
respectively. For daily output files, the time unit is in days since a
specified origin date-time (e.g., days since 1850-01-01 00:00:00)
\citep{taylor2010cmip5}. All CMIP5 output files are required to include
certain metadata \citep{taylor2010cmip5}. This required metadata
includes the experiment, forcing agents input to the model to create the
simulation, time step, institution and institutional contact
information, climate model, and modeling realm \citep{taylor2010cmip5}.
The metadata also must include units for all of the coordinate variables
(e.g., longitude, latitude, time).

\begin{widefigure}
\begin{center}
\includegraphics[width = \textwidth]{netcdf_figure}
\end{center}
\caption{Example of structure of a NetCDF climate model output file for a variable reported at a single depth, like surface air temperature. Data are stored in a three-dimensional array, with measurements at each time step and grid location. Surface temperature data are typically indexed in climate model output files by longitude, latitude, and time, in that order. For example, if the air surface temperature ("tas") is read into an R object called \code{tas}, you can access the value for the first day at the fourth longitude and third latitude with \code{tas[4, 3, 1]}. In addition to the output variable (temperature in this example), vectors with the ordered values of each dimension (longitude, latitude, and time) can also be read in from the netCDF file, as well as attribute data (e.g., units for variables, the calendar used for time).}
\label{fig:netcdfexample}
\end{widefigure}

\citet{taylor2012overview} and \citet{meehl2007wcrp} are excellent
resources for finding out more about the CMIP climate model output data.

@@

A few R packages can be used to work with the netCDF file format used
for CMIP5 files. Earlier packages to work with netCDF files included
\pkg{ncdf} and \pkg{ncvar}, but these do not work with the newer netCDF
version 4 released in 2008 and are no longer available through CRAN.
More recent packages, including \pkg{ncdf4} \citep{ncdf4} and
\pkg{RNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both version
4 and netCDF's older version 3. Climate model output data for CMIP5 is
required to conform with the earlier version (version 3)
\citep{taylor2010cmip5} and so should work with any of these packages,
although it is safer to write code using functions that can be used with
either version in case future phases of CMIP do not require files to
conform with netCDF version 3.

The netCDF format allows you to access metadata and variables describing
the dimensions of the data without reading the full file into memory.
The metadata describes the dimensions of the array and the units, etc.,
of each variable and is printed by the print method of the R object
returned by the \code{nc\_open} function of \pkg{ncdf4}. For CMIP5
files, the dimension variables include the times and locations
corresponding to each array in the data. Before reading in a variable
from a netCDF file, a connection to the file must be opened, for example
with the \code{nc\_open} function from \pkg{ncdf4}. Variables can then
be read in using the \code{ncvar\_get} function from \pkg{ncdf4}, with
the \code{varid} parameter set to ``lat'', ``lon'', or ``time'' (as a
caveat, many climate models output to non-Gregorian calendars, in which
case the time variable should be read in using a different function, as
discussed later in this section). The climate output variable (e.g.,
near-surface air temperature) can similarly be read in using
\code{ncvar\_get}. In this case, the \code{varid} parameter should be
set using the appropriate CMIP5 variable name (e.g., ``tas'' for
near-surface air temperature); these variable names can be found in the
CMIP Requested Output tables \citep{taylor2010cmip5}. You can use the
dimensional time and location data to identify the location of the
variable data you need in the netCDF array and use indexing to read only
that data into memory, without needing to read in the full file (for
example, with the \code{nc.get.var.subset.by.axes} function in
\pkg{ncdf4.helpers}). Once the user is done reading in data from the
file, the connection can be closed (e.g., with the \code{nc\_close}
function from \pkg{ncdf4}).

Since the late 1500s, Western dates have been set using the Gregorian
calendar, which has 365.2425-day years. Some climate models, however,
are run using different calendars, including the Julian calendar
(365.25-day years), a calendar where there are no leap years (365-day
years), a calendar where every year is a leap year (366-day years), and
a calendar of twelve 30-day months (360-day years)
\citep{cfconventions}. With these non-Gregorian calendars, R's base
functions for converting a vector to a Date class based on the number of
days since an origin date (\code{as.Date}, \code{as.POSIXct}) do not
return the desired values. The \pkg{PCICt} \citep{PCICt} and
\pkg{ncdf4.helpers} \citep{ncdf4.helpers} packages provide further
functionality with netCDF files that can be useful when working with
climate model output data and provide particular help in working with
different calendars. CMIP5 netCDF files include information on the
calendar used, which the \code{nc.get.time.series} function in
\pkg{ncdf4.helpers} use to convert the ``time'' variable in the file to
an object of the PCICt class, which provides Date-like functionality for
360- and 365-day calendars \citep{PCICt}. As a note, while these
functions will help with handling most CMIP5 files, the CMIP5 standards
allows use of other calendars which may not be succesfully handled by
these functions, so it is important to assess whether the time variable
range in the PCICt object correctly matches the expected date ranges for
a file as you process CMIP5 data in R.

The size of the climate model output files can be large enough that it
may make more sense to work with smaller chunks of the data, rather than
reading all data into memory and working with the data all at once
\citep{RCMIP5}. This problem aggregates when working with multiple
climate models and more than one ensemble member for each of those
climate models.

The functions and packages described can be used to work with CMIP5
netCDF files in R to do things like map near-surface air temperatures
from a single climate model on a specific day in July 2075 (Figure
\ref{fig:worldmap}) and pull a time series of daily near-surface air
temperature simulations for the climate model grid point closest to
Beijing, China (Figure \ref{fig:timeseries}). The
``starting\_from\_netcdf'' vignette that comes with the
``futureheatwaves'' package provides all code required to create these
figures, as well as more details and code examples on working with CMIP5
netCDF files in R.

@@

@@

The \pkg{RCMIP5} package includes functions to read in CMIP5 data from
netCDF files, scan a directory of CMIP5 files and determine models with
continuous available data, create objects of a special ``cmip5data''
class to work with CMIP5 data within R, and parse the filenames for all
files in a directory to extract information within the filename. For
this package, most functions only work with monthly or less frequent
data (e.g., the functions \code{checkTimePeriod} and \code{cmip5data},
as well as functions that work with \code{cmip5data} objects including
\code{filterDimensions} and \code{getProjectionMatrix}) \citep{RCMIP5}.
While the \code{loadCMIP5} function does successfully load the daily
data as a \code{cmip5data} object, most of the methods for this object
type do not do anything meaningful for daily data. The package's
\code{getFileInfo} function, however, will work with CMIP5 files of any
time step; this function identifies all CMIP5 files in a directory and
creates a dataframe that parses the information contained in the file
name. As a note, the \code{get.split.filename.cmip5} function in the
\pkg{ncdf4.helper} package similarly can be used to parse information
contained in CMIP5 file names \citep{ncdf4.helpers}.

The \pkg{wux} package \citep{wux} includes functions that allow the user
to download CMIP5 monthly-aggregated output directly from within R with
the \code{CMIP5fromESGF} function. However, this function does not allow
downloading of climate model output with finer time steps, like daily
data. This package uses the \code{models2wux} function to read in
climate model output netCDF files and convert to a WUX dataframe that
can be used by other functions in the package. While this function can
input climate model output with daily time steps, if the element
``what.timesteps'' of the \code{modelinput} list input is set to
``daily'', the function aggregates this data to a monthly or less
frequent (e.g., seasonal) aggregation when creating the WUX dataframe.
Therefore, while this package provides very useful functionality for
working with averaged output of daily climate model output data, it
cannot easily be used to identify and characterize multi-day extreme
events like heat waves.

More tips on working with CMIP5 files in R are provided in the
``starting\_from\_netcdf'' vignette of the \pkg{futureheatwaves}
package.

@@

We created the \pkg{futureheatwaves} package to aid in identifying,
characterizing, and exploring multi-day extreme events in daily climate
model output data. Figure \ref{fig:overview} gives an overview of the
two primary functions of the \pkg{futureheatwaves} package. First, the
\code{gen\_hw\_set} function processes a directory of climate projection
files that are stored locally on the user's computer (Figure
\ref{fig:overview}, ``Climate projections''), to generate a list of all
extreme events in each projection, as well as over a dozen
characteristics of each identified extreme event (Table
\ref{tab:hwcharacteristics}). The extreme events are identified and
characterized at one or more study locations (e.g., cities), with study
locations specified by the user in an input file. The extreme events
identified for each ensemble member are output as separate files in a
directory specified by the user (Figure \ref{fig:overview}, ``Extreme
events datasets``).

>>>>

\abstract{%
Research on climate change impacts can require extensive processing,
especially when using ensemble techniques to incorporate output from
different climate models and different simulations of each model. This
processing can be particularly extensive when identifying and
characterizing multi-day extreme events like heat waves and frost day
spells, as these must be processed from model output with daily time
steps. Further, climate model output is in a format and follows
standards that may be unfamiliar to most R users. Here, we provide an
overview of working with daily climate model output data in R. We then
present the \pkg{futureheatwaves} package, which we developed to ease
the process of identifying, characterizing, and exploring multi-day
extreme events in climate model output.
}

@@

Research on climate change impacts can require extensive processing of
climate model output data. This is not only because output files for a
single climate model can be large, but also because of the rising
popularity of ensemble techniques \citep{IPCCch1} in which, to better
characterize uncertainty in projections, impacts are assessed for
multiple climate models, multiple simulations of each climate model, and
multiple climate experiments. Such ensemble techniques help characterize
uncertainty in projections of regional climate change over the next
century due to three distinct sources: (1) internal climate variability,
i.e.~climate noise, (2) climate model uncertainty, i.e.~the same forcing
can produce a different response in different models and (3) scenario
uncertainty, i.e.~uncertainty in future climate forcings (e.g.
\citet{hawkins2009potential}).

A key source of data for ensemble techniques is the Coupled Model
Intercomparison Project, phase 5 (CMIP5; \citet{taylor2012overview}).
This project brought together dozens of major climate modeling groups
around the world to simulate the same future radiative forcing
scenarios, but with their own models. This created an ensemble of
state-of-the-art climate model projections that allow researchers to
study projections and their uncertainties. Most of these modeling groups
additionally performed more than one simulation for each scenario and
model (i.e.~multiple ensemble members), perturbing the initial
conditions by a very tiny amount to quantify uncertainties due to
internal climate variability.

We begin this article with an overview of CMIP5 climate model output
data for R users, focusing on output with a daily time step. We outline
where data from CMIP5 can be obtained as well as how to work with the
file format (netCDF) from R. We overview some R packages that can be
useful when working with this data, as well as aspects of the data
(e.g., non-standard calendars) of which users should be aware when
working with daily climate model output in R.

After this overview, we present the \pkg{futureheatwaves} package, which
we created to aid in identifying and characterizing any type of
multi-day extreme event from daily climate model output (Table
\ref{tab:goals}). The impacts of multi-day extreme events must be
assessed using output in daily time step, unlike other climate impacts
that can be assessed using climate model output at monthly, seasonal, or
yearly time steps. Further, extreme events are identified based on
conditions that are rare for a certain location (e.g., 98th percentile
of local temperature distribution for identifying heat waves)
\citep{IPCCch1}. In this case, assessment is further complicated, as the
event definition must be determined at each study location from climate
model output. Finally, it is often of interest to create summaries of
multiple characteristics of these extreme events. For example, one may
be interested in determining whether the frequency or characteristics
(e.g., length, intensity) of heat waves or warm spells will change under
certain climate change scenarios \citep{IPCCch1}.

The \pkg{futureheatwaves} package handles these challenges and can be
used to identify and characterize a variety of multi-day extreme events
across different ensemble members of one or more climate models. It also
provides some functionality particularly useful in identifying and
characterizing heat waves specifically. Quantification of the impacts of
heat waves on human health suffer from additional sources of uncertainty
beyond those inherent in projections of regional changes in surface
temperature. These include: (1) uncertainty in the definition of a heat
wave itself and (2) uncertainty in the ability of communities to adapt
to changing temperatures (the adaptation scenario). This package
therefore allows the user to create and use a custom extreme event
definition to identify events in the climate model output, as well as
options to explore different scenarios of adaptation to heat.

\begin{table}[t]
\begin{center}
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\begin{tabular}{lL{\dimexpr0.8\textwidth-2\tabcolsep\relax}}
\toprule
 & Design goals of the \pkg{futureheatwaves} package \\
\midrule
1 & Make processing of large ensembles of climate simulations more practical for researchers exploring the potential impacts of heat waves and other multi-day extreme events.\\
2 & Speed up processing time by incorporating C++ in event identification.\\
3 & Keep track of the names of climate models and number of ensemble members processed for each.\\
4 & Not only identify, but also characterize, all extreme events within each climate simulation, to allow the exploration of patterns in these characteristics across different simulations and also to allow the use of more complex impacts models, including models that incorporate event characteristics (e.g., event length, event intensity). For example, this package allows the user to apply a health effects model where risk of mortality is not the same for every heat wave, but rather is modified by heat wave length, intensity, or other measured characteristics.\\
5 & Give users extensive power in customizing the process, including allowing custom extreme event definitions.\\
6 & Allow users to easily explore the extreme events identified within all climate simulations.\\
7 & Create output that is in a ``tidy" data format, allowing it to work well with \pkg{ggplot2} for visualization.\\
\bottomrule
\end{tabular}
\end{center}
\caption{Design goals for the \pkg{futureheatwaves} package.}
\label{tab:goals}
\end{table}

@@

For climate impact studies, a main source for climate model output is
the Coupled Model Intercomparison Project, which is currently in its
fifth phase (CMIP5). Over 20 climate modeling groups created one or more
climate models which, for this project, were run using standardized
scenarios \citep{taylor2012overview}. The resulting output is uniform
across modeling groups and has a consistent structure, which allows
comparison of simulations from different models \citep{IPCCch9}. CMIP5
climate model output is archived at a number of different time steps
(e.g., daily, monthly, seasonal, yearly) \citep{taylor2010cmip5}, and
some variables are reported at multiple levels in the ocean or
atmosphere (e.g., ocean temperature). Here, we will focus on data with a
daily time step for variables reported at a single level (e.g.,
near-surface air temperature).

Each modeling group ran simulations under several experiments, with
experiments varying in terms of radiative forcing by using different
scenarios of time-varying model inputs (greenhouse gas emissions or
concentrations, land use changes, etc.)
\citep{taylor2012overview, IPCCch9}. Experiments include historical
experiments (run using radiative forcing consistent with observed and
reconstructed data for 1850--2005), pre-Industrial control experiments,
and experiments of future scenarios of radiative forcing over the 21st
century or longer (e.g., RCP4.5, RCP8.5) \citep{taylor2012overview}.
Some modeling groups created ensembles of output for a specific model
and experiment, in which they ran the experiment multiple times with the
model with very small changes to the initial conditions.

The CMIP5 climate model output data is distributed across data nodes at
different climate modeling centers \citep{taylor2012overview}, but can
be accessed centrally at the World Climate Research Programme CMIP5 data
portal at \url{https://pcmdi.llnl.gov/search/cmip5/}. Users must
register before downloading data, and some data are restricted to
non-commercial use. There is a separate file for each combination of
climate model, experiment, modeling realm (e.g., atmosphere, ocean),
variable, time step, and ensemble member
\citep{taylor2012overview, taylor2010cmip5}. For finer time scales, the
output is further split across multiple files for specific year ranges
(e.g., 5 years of output for each file) \citep{taylor2010cmip5}. Each
file's name includes the output variable, climate model, experiment, and
ensemble member for the simulation \citep{taylor2010cmip5}.

CMIP5 files can be searched and downloaded through a point-and-click web
interface. They can also be downloaded in bulk to computers with Unix or
Mac operating systems using the \code{wget} file downloading utility.
Appropriate \code{wget} scripts can be created either through the World
Climate Research Programme CMIP5 data portal or through Earth System
Grid Federation's Search RESTful API. Tips on efficiently searching and
downloading the data, including through use of wget scripts and the
search API, are available as user tutorials through the website of the
University of Colorado Boulder's Earth System CoG (e.g.,
\url{https://www.earthsystemcog.org/projects/cog/doc/wget} for a
tutorial on downloading files using wget).

CMIP5 files are saved in Network Common Data Format (netCDF), a binary
file format that allows storage of data representing a regular array.
For climate model output at a single level (e.g., near-surface air
temperature), the data is a 3-dimensional array, with dimensions
representing time and two coordinates of location (e.g., latitude and
longitude). Figure \ref{fig:netcdfexample} provides a sketch of the
structure of netCDF files for single-level climate model output. Global
climate models generate output at regularly-spaced time steps, typically
at regularly-spaced grid points around the world. The latitude and
longitude spacing of grid points vary by climate model, but are
typically 1--2 degrees for atmospheric variables in CMIP5 models
\citep{IPCCch9}. Each data point in the netCDF array gives the modeled
value of the variable (e.g., surface temperature) for a single time
point and location. For CMIP5 climate model output, the location units
are in degrees east and degrees north for longitude and latitude,
respectively. For daily output files, the time unit is in days since a
specified origin date-time (e.g., days since 1850-01-01 00:00:00)
\citep{taylor2010cmip5}. All CMIP5 output files are required to include
certain metadata \citep{taylor2010cmip5}, including the experiment,
forcing agents input to the model to create the simulation, time step,
institution and institutional contact information, climate model, and
modeling realm \citep{taylor2010cmip5}. The metadata also must include
units for all of the coordinate variables (e.g., longitude, latitude,
time). The netCDF format allows you to access metadata and variables
describing the dimensions of the data without reading the full file into
memory.

\begin{figure}
\begin{center}
\includegraphics[width = \textwidth]{netcdf_figure}
\end{center}
\caption{Example of structure of a NetCDF climate model output file for a variable reported at a single depth, like surface air temperature. Data are stored in a three-dimensional array, with measurements at each time step and grid location. Surface temperature data are typically indexed in climate model output files by longitude, latitude, and time, in that order. For example, if the air surface temperature ("tas") is read into an R object called \code{tas}, you can access the value for the first day at the fourth longitude and third latitude with \code{tas[4, 3, 1]}. In addition to the output variable (temperature in this example), vectors with the ordered values of each dimension (longitude, latitude, and time) can also be read in from the netCDF file, as well as attribute data (e.g., units for variables, the calendar used for time).}
\label{fig:netcdfexample}
\end{figure}

To find out more about the CMIP climate model output data,
\citet{taylor2012overview} and \citet{meehl2007wcrp} are excellent
resources.

@@

When working with daily climate model output data, challenges to R users
include: (1) the file format; (2) use of non-Gregorian calendars; and
(3) large file sizes. This section explains these challenges and offers
some strategies for dealing with them.

@@

CMIP5 data are available as netCDF files, for which base R import
functions do not exist. There are, however, a few R packages that can be
used to work with the netCDF file format used for CMIP5 files. Older
packages include \pkg{ncdf} and \pkg{ncvar}, but these do not work with
the newer netCDF version 4 released in 2008 and are no longer available
through CRAN. More recent packages, including \pkg{ncdf4} \citep{ncdf4}
and \pkg{RNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both
version 4 and netCDF's older version 3. While climate model output data
for CMIP5 is required to conform with the earlier version (version 3)
\citep{taylor2010cmip5}, it is safer to write code using functions that
can be used with either version, in case future phases of CMIP do not
require files to conform with netCDF version 3.

You can do a number of things with netCDF files in R using these
functions. For example, \pkg{ncdf4}'s \code{nc\_open} function can be
used to open a connection to a netCDF file, and the object returned by
the function includes the file's attribute data. Once a file connection
is open, variables can be read in using the \code{ncvar\_get} function.
For example, the variables defining the dimensions of the sketched
netCDF file in Figure \ref{fig:netcdfexample} could be read into R with
\code{ncvar\_get} with the \code{varid} parameter set to ``lat'',
``lon'', or ``time''. The climate output variable (e.g., near-surface
air temperature) can similarly be read in using \code{ncvar\_get}. In
this case, the \code{varid} parameter should be set using the
appropriate CMIP5 variable name (e.g., ``tas'' for near-surface air
temperature); these variable names can be found in the CMIP Requested
Output tables \citep{taylor2010cmip5}. If you only need data from a
subset of the full file, you can use the dimensional time and location
data to identify the location of the variable data you need in the
netCDF array and use indexing to read only that data into memory,
without needing to read in the full file (for example, with the
\code{nc.get.var.subset.by.axes} function in \pkg{ncdf4.helpers}). Once
the user is done reading in data from the file, the connection can be
closed (e.g., with the \code{nc\_close} function from \pkg{ncdf4}).

A second challenge when working with climate model output data in R is
that some climate models output to non-Gregorian calendars. Since the
late 1500s, Western dates have been set using the Gregorian calendar,
which has 365.2425-day years. Some climate models, however, are run
using different calendars, including the Julian calendar (365.25-day
years), a calendar where there are no leap years (365-day years), a
calendar where every year is a leap year (366-day years), and a calendar
of twelve 30-day months (360-day years) \citep{cfconventions}. With
these non-Gregorian calendars, R's base functions for converting a
vector to a Date class based on the number of days since an origin date
(\code{as.Date}, \code{as.POSIXct}) do not return the desired values.

Two R packages provide help with non-Gregorian calendars: \pkg{PCICt}
\citep{PCICt} and \pkg{ncdf4.helpers} \citep{ncdf4.helpers}. The
\code{nc.get.time.series} function in \pkg{ncdf4.helpers} pulls and uses
metadata on the calendar stored in the CMIP5 netCDF file's attributes to
convert the ``time'' variable in the file to an object of the
\code{PCICt} class. This class is defined in the \pkg{PCICt} package and
provides Date-like functionality for 360- and 365-day calendars
\citep{PCICt}. However, while these functions will help with handling
most CMIP5 files, the CMIP5 standards allows use of other calendars
which may not be succesfully handled by these functions, so it is
important to assess whether the time variable range in the \code{PCICt}
object correctly matches the expected date ranges for a file when
processing CMIP5 data in R.

Finally, the size of CMIP5 files can make them difficult to work with in
R. CMIP5 climate model output files can be as large as several
gigabytes. The size of the files can therefore be large enough that it
may make more sense to work with smaller chunks of the data in R, rather
than reading all data into memory and working with the data all at once
\citep{RCMIP5}. This problem aggregates when working with multiple
climate models and more than one ensemble member for each of those
climate models.

@@

The \pkg{RCMIP5} package includes functions to read in CMIP5 data from
netCDF files, scan a directory of CMIP5 files and determine models with
continuous available data, create objects of a special \code{cmip5data}
class to work with CMIP5 data within R, and parse the filenames for all
files in a directory to extract information within the filename. For
this package, most functions only work with monthly or less frequent
data \citep{RCMIP5}. While the \code{loadCMIP5} function does
successfully load daily data as a \code{cmip5data} object, most of the
methods for this object type do not do anything meaningful for daily
data. The package's \code{getFileInfo} function, however, will work with
CMIP5 files of any time step; this function identifies all CMIP5 files
in a directory and creates a dataframe with information parsed from the
file name. As a note, the \code{get.split.filename.cmip5} function in
the \pkg{ncdf4.helper} package similarly can be used to parse
information contained in CMIP5 file names \citep{ncdf4.helpers}.

The \pkg{wux} package \citep{wux} includes functions that allow the user
to download CMIP5 monthly-aggregated output directly from within R with
the \code{CMIP5fromESGF} function. The package then uses the
\code{models2wux} function to read in climate model output netCDF files
and convert them to ``WUX'' dataframes, which can be used by other
functions in the package. While this function can input climate model
output with daily time steps (the ``what.timesteps'' element of the
\code{modelinput} list input must be set to ``daily''), the function
aggregates this data to a monthly or less frequent (e.g., seasonal)
aggregation when creating the WUX dataframe. Therefore, while this
package provides very useful functionality for working with averaged
output of daily climate model output data, it cannot easily be used to
identify and characterize multi-day extreme events like heat waves.

The functions and packages described in this section can be used with
CMIP5 netCDF files to do things in R like map near-surface air
temperatures from a single climate model on a specific day (Figure
\ref{fig:worldmap}) or pull a time series of daily near-surface air
temperature simulations at a specific climate model grid point (Figure
\ref{fig:timeseries}). The ``starting\_from\_netcdf'' vignette that
comes with the ``futureheatwaves'' package provides all code required to
create these figures, as well as more details and code examples on
working with CMIP5 netCDF files in R.

@@

We created the \pkg{futureheatwaves} package to aid in identifying,
characterizing, and exploring multi-day extreme events in daily climate
model output data. Figure \ref{fig:overview} gives an overview of the
two primary functions of the \pkg{futureheatwaves} package. First, the
\code{gen\_hw\_set} function processes a directory of climate projection
files that are stored locally on the user's computer (Figure
\ref{fig:overview}, ``Climate projections''), to generate a list of all
extreme events in each projection, as well as over a dozen
characteristics of each identified extreme event (Table
\ref{tab:hwcharacteristics}). This package start from files rather than
R objects to avoid loading data from all climate model ensembles at
once; instead, the function loads, processes, and saves output for a
single climate model ensemble member at a time. The extreme events are
identified and characterized at one or more study locations (e.g.,
cities), which the user specifies in an input file. The extreme events
identified for each ensemble member are output as separate files in a
directory specified by the user (Figure \ref{fig:overview}, ``Extreme
events datasets``).

Once the user creates these data frames of location-specific heat waves,
the \code{apply\_all\_models} function can be used to apply custom
functions across all the extreme event data frames generated by
\code{gen\_hw\_set}. This functionality allows users to create summaries
of extreme events across all climate models and ensemble members (Figure
\ref{fig:overview}, right). The function can be used to generate summary
statistics (e.g., determine average heat wave length or total frost
days) or to apply more complex functions (e.g., apply epidemiologic
effect estimates across the heat waves to generate health impact
estimates).
